Supplemental Material for ``Assessing United States county-level exposure for research on tropical cyclones and human health"

### 2019-07-29 writing/ehp_discussion.tex
Update figure references

We investigated how well these data correspond with data from other available
sources (Figures~\ref{fig:windcomparison}\,--\,\ref{fig:floodcomparison}
and~S6\,--\,7). While generally in agreement with data from other sources,
there are a few caveats that should be considered when applying these data in
future studies. The rainfall data are from re-analysis data are generally
well-correlated with observed ground-based station data, but may sometimes
oversmooth very high rainfall values compared to ground-based observations
(Figure~\ref{fig:raincomparison}). When rainfall data is used to create binary
exposure classifications, this disagreement is unlikely to influence results,
as both data sources agree in identifying these as storms with high rainfall,
but it may be important to consider for cases that include rainfall as a
continuous measurement.

The wind data are based on modeled, rather than observed, values, and while the
modeled wind data generally agree well with post-analysis maximum wind radii
from \ac{HURDAT2}~\parencite{landsea2013} (Figure~\ref{fig:windcomparison}),
there were a few storms where the two datasets did not agree as well ($<$90\%
of counties classified into the same wind categories by the two sources of wind
estimates). These storms were exceptionally large systems for which high winds
persisted well inland from landfall, like Hurricanes Sandy and Ike
(Figures~\ref{fig:windcomparison} and\~S6). Since the wind radii are based on
\ldots, they have [advantages]; their main disadvantages is that they provide
categorical, rather than continuous, estimates of wind speed, and in many cases
researchers might prefer more precise estimates of storm wind speeds. However,
since the wind radii data may be useful in some cases as either a primary
metric or for sensitivity analysis, we have include county-level wind estimates
for each storm based on these wind radii, in addition to the continuous modeled
windspeeds, in the open-source data \cite{hurricaneexposuredata}.

For the flooding events data, we found it often corresponded with measurements
from \ac{US} Geographical Survey streamgages, with flood event listings more
likely during storms which elevated stream flows. However, there are
differences between the two flooding datasets, and these highlight both the
difficulty of measuring flood exposure at the county level and inherent
challenges in using the storm event database for epidemiologic exposure
assessment.  The flood and tornado data came from the \ac{NOAA} Storm Events
database, which, while a widely-used database of events maintained by
\ac{NOAA}, is based on reports, and so may be prone to
underreporting~\parencite{Ashley2008flood, Curran2000}, especially in less
populated areas~\parencite{Witt1998, Ashley2007}, as well as to other reporting
errors.

@@

We found geographic patterns in tropical cyclones based on the different
exposure metrics that. These patterns were not unexpected based on what is
known about tropical cyclone hazards, but still highlight variations across
metrics that are critical to consider in designing studies for tropical cyclone
epidemiology, and they demonstrate the need for multi-hazard exposure datasets
for tropical cyclone epidemiology. Wind-based exposures had a strong coastal
pattern, which is consistent with the dramatic decrease in wind intensity that
typically characterizes the landfall of tropical cyclones. Exposures by rain-
and distance-based metrics often extended further inland compared to the
wind-based metric, up to the Appalachian mountains. This agrees with previous
research indicating that the Appalachian mountains' topography both enhances
precipitation during tropical cyclones and provides hydrological conditions for
severe flooding~\parencite{rees2001}.  Almost all tornado-based exposures were
in southern coastal states, echoing previous findings that most noteworthy
tropical cyclone-related tornadoes occur on the right-side of the tropical
cyclone track in Atlantic-basin \ac{US} storms~\parencite{moore2012}.  These
exposure averages are somewhat limited as estimates of long-term frequencies,
as tropical cyclones follow decadal patterns~\parencite{kossin2007more} likely
not adequately captured in the available data. However, exposure assessments
spanning over a century have found similar patterns for distance- and
wind-based metrics~\textcite{zandbergen2009, kruk2010} [double-check that refs
found similar patterns].

@@

We found that there was typically only moderate or low agreement between
county-level exposure classifications for a tropical cyclone when comparing
different exposure metrics. This suggests that the use of one of these metrics
as a proxy for all storm-based hazards in epidemiological research would often
be problematic, introducing exposure misclassification.

@@

However, we did find a small set of tropical cyclones for which for which
agreement between metrics were unusually high (e.g., Floyd in~1999, Irene in
2011, Hannah in~2008, Bertha in~1996).  Hurricanes Floyd in~1999 and Irene
in~2011 both made their first \ac{US} landfall in North Carolina at minor
hurricane windspeeds (Category~2 and~1, respectively) and then skimmed the
eastern coast of the \ac{US} north through New England, bringing large rainfall
to much of the eastern coast from North Carolina north and causing extensive
inland flooding in North Carolina (Floyd) and New England
(Irene)~\parencite{avila2013atlantic, lawrence2000atlantic}.  Hurricanes Hannah
in~2008 and Bertha in~1996 followed the eastern coastline north of North
Carolina in a pattern similar to Floyd and Irene. For this set of storms, the
tropical cyclones' persistent proximity to water may have helped maintain wind
speeds in similar patterns to rain and distance exposures, resulting in this
moderate to good agreement among exposure assessments based on different
metrics.

>>>>

We investigated how well these data correspond with data from other available
sources (Figures~\ref{fig:raincomparison}\,--\,\ref{fig:floodcomparison}
and~S2). While generally in agreement with data from other sources,
there are a few caveats that should be considered when applying these data in
future studies. The rainfall data are from re-analysis data are generally
well-correlated with observed ground-based station data, but may sometimes
oversmooth very high rainfall values compared to ground-based observations
(Figure~\ref{fig:raincomparison}). When rainfall data is used to create binary
exposure classifications, this disagreement is unlikely to influence results,
as both data sources agree in identifying these as storms with high rainfall,
but it may be important to consider for cases that include rainfall as a
continuous measurement.

The wind data are based on modeled, rather than observed, values, and while the
modeled wind data generally agree well with post-analysis maximum wind radii
from \ac{HURDAT2}~\parencite{landsea2013} (Figure~\ref{fig:windcomparison}),
there were a few storms where the two datasets did not agree as well ($<$90\%
of counties classified into the same wind categories by the two sources of wind
estimates). These storms were exceptionally large systems for which high winds
persisted well inland from landfall, like Hurricanes Sandy and Ike
(Figures~\ref{fig:windcomparison} and~S2). Since the wind radii are based on
\ldots, they have [advantages]; their main disadvantages is that they provide
categorical, rather than continuous, estimates of wind speed, and in many cases
researchers might prefer more precise estimates of storm wind speeds. However,
since the wind radii data may be useful in some cases as either a primary
metric or for sensitivity analysis, we have include county-level wind estimates
for each storm based on these wind radii, in addition to the continuous modeled
windspeeds, in the open-source data \parencite{hurricaneexposuredata}.

For the flooding events data, we found it often corresponded with measurements
from \ac{US} Geographical Survey streamgages, with flood event listings more
likely during storms which elevated stream flows
(Figure~\ref{fig:floodcomparison}. However, there are differences between the
two flooding datasets, and these highlight both the difficulty of measuring
flood exposure at the county level and inherent challenges in using the storm
event database for epidemiologic exposure assessment.  The flood and tornado
data came from the \ac{NOAA} Storm Events database, which, while a widely-used
database of events maintained by \ac{NOAA}, is based on reports, and so may be
prone to underreporting~\parencite{Ashley2008flood, Curran2000}, especially in
less populated areas~\parencite{Witt1998, Ashley2007}, as well as to other
reporting errors.

@@

We found geographic patterns in tropical cyclones based on the different
exposure metrics (Figure~\ref{fig:averageexposures}). These patterns were not
unexpected based on what is known about tropical cyclone hazards, but still
highlight variations across metrics that are critical to consider in designing
studies for tropical cyclone epidemiology, and they demonstrate the need for
multi-hazard exposure datasets for tropical cyclone epidemiology. Wind-based
exposures had a strong coastal pattern, which is consistent with the dramatic
decrease in wind intensity that typically characterizes the landfall of
tropical cyclones. Exposures by rain- and distance-based metrics often extended
further inland compared to the wind-based metric, up to the Appalachian
mountains. This agrees with previous research indicating that the Appalachian
mountains' topography both enhances precipitation during tropical cyclones and
provides hydrological conditions for severe flooding~\parencite{rees2001}.
Almost all tornado-based exposures were in southern coastal states, echoing
previous findings that most noteworthy tropical cyclone-related tornadoes occur
on the right-side of the tropical cyclone track in Atlantic-basin \ac{US}
storms~\parencite{moore2012}.  These exposure averages are somewhat limited as
estimates of long-term frequencies, as tropical cyclones follow decadal
patterns~\parencite{kossin2007more} likely not adequately captured in the
available data. However, exposure assessments spanning over a century have
found similar patterns for distance- and wind-based
metrics~\textcite{zandbergen2009, kruk2010} [double-check that refs found
similar patterns].

@@

We found that there was typically only moderate or low agreement between
county-level exposure classifications for a tropical cyclone when comparing
different exposure metrics
(Figures~\ref{fig:ivanexposure}\,--\,\ref{fig:jaccard}  and S4\,--\,S5). This
suggests that the use of one of these metrics as a proxy for all storm-based
hazards in epidemiological research would often be problematic, introducing
exposure misclassification.

@@

However, we did find a small set of tropical cyclones for which for which
agreement between metrics were unusually high (e.g., Floyd in~1999, Irene in
2011, Hannah in~2008, Bertha in~1996; Figure~\ref{fig:jaccard}).  Hurricanes
Floyd in~1999 (Figure~S4) and Irene in~2011 both made their first \ac{US}
landfall in North Carolina at minor hurricane windspeeds (Category~2 and~1,
respectively) and then skimmed the eastern coast of the \ac{US} north through
New England, bringing large rainfall to much of the eastern coast from North
Carolina north and causing extensive inland flooding in North Carolina (Floyd)
and New England (Irene)~\parencite{avila2013atlantic, lawrence2000atlantic}.
Hurricanes Hannah in~2008 and Bertha in~1996 followed the eastern coastline
north of North Carolina in a pattern similar to Floyd and Irene. For this set
of storms, the tropical cyclones' persistent proximity to water may have helped
maintain wind speeds in similar patterns to rain and distance exposures,
resulting in this moderate to good agreement among exposure assessments based
on different metrics.

### 2019-07-29 writing/ehp_methods.tex
Update figure references

Two of the exposure metrics (flood- and tornado-based) were inherently binary,
since these metrics were based on whether an event was listed in the NOAA Storm
Events database.  For the other exposure metrics, each county was classified as
exposed to a tropical cyclone based on whether the exposure metric exceeded a
certain threshold (Table~\ref{tab:exposuremetrics}). For the rainfall metric, a
distance constraint was also necessary, to ensure that rains unrelated and far
from the storm track were not misattributed to a storm. Through exploratory
analysis, we set this distance metric at~500 \si{\kilo\metre} (i.e., for a
county to be classified as exposed based on rainfall, the cumulative rainfall
had to be over 75 \si{\milli\metre} and the storm must have passed within~500
\si{\kilo\metre} of the county; Table~\ref{tab:exposuremetrics}). We set this
distance constraint at a value that was typically large enough to capture
storm-related rain. However, data users should note that in rare examples of
exceptionally large storms (e.g., Hurricane Ike in 2008) or storms for which
storm tracking was stopped at extratropical transition (e.g., Tropical Storm
Lee in 2011), some storm-related rains may be missed because of this distance
constraint (Figure~S7).

>>>>

Two of the exposure metrics (flood- and tornado-based) were inherently binary,
since these metrics were based on whether an event was listed in the NOAA Storm
Events database.  For the other exposure metrics, each county was classified as
exposed to a tropical cyclone based on whether the exposure metric exceeded a
certain threshold (Table~\ref{tab:exposuremetrics}). For the rainfall metric, a
distance constraint was also necessary, to ensure that rains unrelated and far
from the storm track were not misattributed to a storm. Through exploratory
analysis, we set this distance metric at~500 \si{\kilo\metre} (i.e., for a
county to be classified as exposed based on rainfall, the cumulative rainfall
had to be over 75 \si{\milli\metre} and the storm must have passed within~500
\si{\kilo\metre} of the county; Table~\ref{tab:exposuremetrics}). We set this
distance constraint at a value that was typically large enough to capture
storm-related rain. However, data users should note that in rare examples of
exceptionally large storms (e.g., Hurricane Ike in 2008) or storms for which
storm tracking was stopped at extratropical transition (e.g., Tropical Storm
Lee in 2011), some storm-related rains may be missed because of this distance
constraint (Figure~S1).

### 2019-07-29 writing/ehp_results.tex
Update figure references

For maximum sustained surface wind speed estimates, we found that the modeled
estimates, which we included as the primary wind exposure metric in the open
source data and used for further analysis in this study, generally agreed well
with estimates based on the wind radii reported in \ac{HURDAT2}. For most
storms, $\ge$95\% of counties were assigned the same category of wind speed
($<$34~\si{\knot}; 34\,--\,49.9~\si{\knot}; 50\,--\,63.9~\si{\knot};
$\ge$64~\si{\knot}) by both data sources (Figure~\ref{fig:windcomparison}).
Disagreement was limited to large storms that were characterized by high winds
well inland, like Hurricanes Sandy in 2012 and Ike in 2008. For these two
storms, the modeled wind speed values somewhat overestimated the extent of the
storm's most severe winds at landfall but then underestimated the extent of
34\,--\,49.9 \si{\knot} storm winds further inland (Figure~S6). For researchers
who would like to conduct sensitivity analysis between these two sets of wind
data, we have included estimates from the \ac{HURDAT2} wind radii as a
secondary measure of wind speeds in the open-source dataset
\parencite{hurricaneexposuredata}.

@@

When we calculated and mapped the average number of exposures per decade in
each county for each exposure metric (Figure~\ref{fig:averageexposure}), strong
geographical patterns were clear. Wind-based exposure had a strong coastal
pattern, with almost all exposures in counties within about~200
\si{\kilo\metre} (124~mi) of the coastline. While exposures by rain- and
distance-based metrics were also more common in coastal areas compared to
inland areas, they also created inland exposures that were not identified by
the wind-based metric. Rain and, to some extent, flood exposures were
characterized by a pattern defined by the Appalachian Mountains, with much
lower exposures west of the mountain range than to the east. Almost all
tornado-based exposures were in coastal states, with most in Florida, Alabama,
South Carolina, and North Carolina and almost none north of Maryland.
Flood-based exposures were highest in North Carolina and along the South
Carolina coast, as well as further north in New Jersey, southeastern
Pennsylvania, and southeastern New York. Notably, some areas of the \ac{US}
that experienced regular exposures to tropical cyclone-related excessive
rainfall and floods were not identified as areas of regular risk based on the
more commonly used tropical cyclone exposure assessment metrics of wind or
distance (Figure~\ref{fig:averageexposure}). Patterns were similar when
analysis was restricted to years with all exposure data available
(1996\,--\,2011; Figure~S5).

@@

Figure~\ref{fig:ivanexposure} shows Hurricane Ivan in 2004 as an example. For
the distance-based metric, the counties assessed as exposed followed the
tropical cyclone's track, including counties well inland. For the wind metric,
only counties near two of the tropical cyclone's landfalls were assessed as
exposed. For rain- and flood-based metrics, however, exposure extended to the
left of the track, including counties as far north as New York and Connecticut,
while for the tornado metric, exposed counties tended to be to the right of the
track and included several counties in central North Carolina, South Carolina,
and Georgia that were not identified as exposed to Ivan based on any other
metric. Figure~S3 provides similar maps for four other example tropical
cyclones (selected because they exposed many \ac{US}  counties based on at
least one metric), and Figure~S4 provides four examples of tropical cyclones in
which the disagreement in exposure assessment between rain- and wind-based
metrics was particularly notable.

@@

\begin{comment}
Storms are displayed within
clusters that have similar patterns in county-level exposure agreement for
metric pairs, based on hierarchical clustering using the complete link
method~\parencite{murtagh2012algorithms} (i.e., storms in the same cluster tend
to have similar patterns for the pairwise strength of agreement among metrics);
columns are also ordered based on hierarchical clustering. 
Maps are
available showing the counties identified as exposed under each of five metrics
for the widest-extent storm in each cluster: Hurricane Ivan in~2004
(Figure~\ref{fig:ivanexposure}) and Hurricanes Floyd in~1999, Lee in~2011, Cindy
in~2005, and Katrina in~2005 (Figure~S3).
\end{comment}

There were, however, a few exceptions\,---\,tropical cyclones in which similar
counties were determined to be exposed to the tropical cyclone for two or more
of the metrics considered.  For Floyd in~1999 (Figure~S3) and Irene in~2011,
for example, county-level classification agreed moderately to well (Jaccard
index of~$\sim 0.5\mbox{\,--\,}0.8$) for all pairs of exposure metrics except
those including the tornado-based metric. For another set of tropical cyclones
(e.g., Lee in~2011 [Figure~S3], Ernesto in~2006, and Bertha in~1996), there was
moderate to good agreement for all pairwise combinations of distance, rain, and
wind, but poor agreement for all other combinations of metrics. However, these
two sets of tropical cyclones represented the minority of all tropical cyclones
considered; for most tropical cyclones, there was low overlap in the counties
determined to be exposed to the tropical cyclones between most pairings of
metrics.

>>>>

For maximum sustained surface wind speed estimates, we found that the modeled
estimates, which we included as the primary wind exposure metric in the open
source data and used for further analysis in this study, generally agreed well
with estimates based on the wind radii reported in \ac{HURDAT2}. For most
storms, $\ge$95\% of counties were assigned the same category of wind speed
($<$34~\si{\knot}; 34\,--\,49.9~\si{\knot}; 50\,--\,63.9~\si{\knot};
$\ge$64~\si{\knot}) by both data sources (Figure~\ref{fig:windcomparison}).
Disagreement was limited to large storms that were characterized by high winds
well inland, like Hurricanes Sandy in 2012 and Ike in 2008. For these two
storms, the modeled wind speed values somewhat overestimated the extent of the
storm's most severe winds at landfall but then underestimated the extent of
34\,--\,49.9 \si{\knot} storm winds further inland (Figure~S2). For researchers
who would like to conduct sensitivity analysis between these two sets of wind
data, we have included estimates from the \ac{HURDAT2} wind radii as a
secondary measure of wind speeds in the open-source dataset
\parencite{hurricaneexposuredata}.

@@

When we calculated and mapped the average number of exposures per decade in
each county for each exposure metric (Figure~\ref{fig:averageexposure}), strong
geographical patterns were clear. Wind-based exposure had a strong coastal
pattern, with almost all exposures in counties within about~200
\si{\kilo\metre} (124~mi) of the coastline. While exposures by rain- and
distance-based metrics were also more common in coastal areas compared to
inland areas, they also created inland exposures that were not identified by
the wind-based metric. Rain and, to some extent, flood exposures were
characterized by a pattern defined by the Appalachian Mountains, with much
lower exposures west of the mountain range than to the east. Almost all
tornado-based exposures were in coastal states, with most in Florida, Alabama,
South Carolina, and North Carolina and almost none north of Maryland.
Flood-based exposures were highest in North Carolina and along the South
Carolina coast, as well as further north in New Jersey, southeastern
Pennsylvania, and southeastern New York. Notably, some areas of the \ac{US}
that experienced regular exposures to tropical cyclone-related excessive
rainfall and floods were not identified as areas of regular risk based on the
more commonly used tropical cyclone exposure assessment metrics of wind or
distance (Figure~\ref{fig:averageexposure}). Patterns were similar when
analysis was restricted to years with all exposure data available
(1996\,--\,2011; Figure~S3).

@@

Figure~\ref{fig:ivanexposure} shows Hurricane Ivan in 2004 as an example. For
the distance-based metric, the counties assessed as exposed followed the
tropical cyclone's track, including counties well inland. For the wind metric,
only counties near two of the tropical cyclone's landfalls were assessed as
exposed. For rain- and flood-based metrics, however, exposure extended to the
left of the track, including counties as far north as New York and Connecticut,
while for the tornado metric, exposed counties tended to be to the right of the
track and included several counties in central North Carolina, South Carolina,
and Georgia that were not identified as exposed to Ivan based on any other
metric. Figure~S4 provides similar maps for four other example tropical
cyclones (selected because they exposed many \ac{US}  counties based on at
least one metric), and Figure~S5 provides four examples of tropical cyclones in
which the disagreement in exposure assessment between rain- and wind-based
metrics was particularly notable.

@@

\begin{comment}
Storms are displayed within
clusters that have similar patterns in county-level exposure agreement for
metric pairs, based on hierarchical clustering using the complete link
method~\parencite{murtagh2012algorithms} (i.e., storms in the same cluster tend
to have similar patterns for the pairwise strength of agreement among metrics);
columns are also ordered based on hierarchical clustering. 
Maps are
available showing the counties identified as exposed under each of five metrics
for the widest-extent storm in each cluster: Hurricane Ivan in~2004
(Figure~\ref{fig:ivanexposure}) and Hurricanes Floyd in~1999, Lee in~2011, Cindy
in~2005, and Katrina in~2005 (Figure~S4).
\end{comment}

There were, however, a few exceptions\,---\,tropical cyclones in which similar
counties were determined to be exposed to the tropical cyclone for two or more
of the metrics considered.  For Floyd in~1999 (Figure~S4) and Irene in~2011,
for example, county-level classification agreed moderately to well (Jaccard
index of~$\sim 0.5\mbox{\,--\,}0.8$) for all pairs of exposure metrics except
those including the tornado-based metric. For another set of tropical cyclones
(e.g., Lee in~2011 [Figure~S3], Ernesto in~2006, and Bertha in~1996), there was
moderate to good agreement for all pairwise combinations of distance, rain, and
wind, but poor agreement for all other combinations of metrics. However, these
two sets of tropical cyclones represented the minority of all tropical cyclones
considered; for most tropical cyclones, there was low overlap in the counties
determined to be exposed to the tropical cyclones between most pairings of
metrics.
